%% The ETH regulations taking effect from November 2013 state that the following additional information must be included in research plans:
%% \begin{itemize}
%%  \item Expected publications
%%  \item Teaching duties
%%  \item Other duties
%% \end{itemize}
%% This section is indented to list any special conditions stipulated by your supervisor such as a publication in a specific conference or journal, or special teaching and administration duties. If there are none, you can simply write ``no special requirements''.

\paragraph{Accepted works}
\begin{itemize}
\item Predicting Circulatory System Deterioration in Intensive Care Unit Patients, \textit{Joint Workshop on Artificial Intelligence in Health 2018}. (Workshop paper, \textbf{WP 1})
\item A Machine Learning-based Early Warning System for Circulatory System Deterioration in Intensive Care Unit Patients, \textit{AMIA 2018 Annual Symposium}. (Podium abstract, \textbf{WP 1})
\end{itemize}
\paragraph{Expected publications}
\begin{itemize}
\item A medical journal paper on ICU Bern organ deterioration early warning system project. (\textbf{WP 1})
\item A conference paper that benchmarks the unsupervised representation learning methods on medical time series, \textit{submitted to NIPS 2018}. (\textbf{WP 2})
\item A conference paper on representation learning with the proposed algorithm that will address the multirate challenges in medical time series. (\textbf{WP 3})
\item A conference paper describing the proposed method to deal with the irregularity issues in medical time series. (\textbf{WP 3})
\item A conference paper presenting the proposed method to handle the missing variable problems in medical time series. (\textbf{WP 4})
\end{itemize}
\paragraph{Teaching duties}
The teaching plan is to TA one course from the machine learning institute of ETH every semester.
\paragraph{Student project supervision}
The student project supervision plan is to supervise one master project maybe along with a short-term student project each year.
So far, I have supervised one master project on spliceogenicity prediction from Aug. 2017 to Feb. 2018, and am currently supervising another one on cancer mutational signatures which started in Jun. 2018.
\paragraph{Other projects}
\begin{itemize}
\item \textbf{Testing for differential protein abundance from DIA data}. The goal of this project is to develop new methods for identifying differentially expressed proteins from the mass spectrometry (MS) data.
A paper on this project will be expected in winter 2018.
\end{itemize}
